# Mathematics and Science {#math-sci}

<!-- Required to number equations in HTML files -->
<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>

## Math

\TeX\ is the best way to typeset mathematics. Donald Knuth designed \TeX\ when he got frustrated at how long it was taking the typesetters to finish his book, which contained a lot of mathematics.  One nice feature of _R Markdown_ is its ability to read LaTeX code directly.

If you are doing a thesis that will involve lots of math, you will want to read the following section which has been commented out. If you're not going to use math, skip over or delete this next commented section.


<!-- MATH and PHYSICS majors: Uncomment the following section -->
<!--
$$\sum_{j=1}^n (\delta\theta_j)^2 \leq {{\beta_i^2}\over{\delta_i^2 + \rho_i^2}}
\left[ 2\rho_i^2 + {\delta_i^2\beta_i^2\over{\delta_i^2 + \rho_i^2}} \right] \equiv \omega_i^2
$$

From Informational Dynamics, we have the following (Dave Braden):

After _n_ such encounters the posterior density for $\theta$ is

$$
\pi(\theta|X_1< y_1,\dots,X_n<y_n) \varpropto \pi(\theta) \prod_{i=1}^n\int_{-\infty}^{y_i}
   \exp\left(-{(x-\theta)^2\over{2\sigma^2}}\right)\ dx
$$

Another equation:

$$\det\left|\,\begin{matrix}%
c_0&c_1\hfill&c_2\hfill&\ldots&c_n\hfill\cr
c_1&c_2\hfill&c_3\hfill&\ldots&c_{n+1}\hfill\cr
c_2&c_3\hfill&c_4\hfill&\ldots&c_{n+2}\hfill\cr
\,\vdots\hfill&\,\vdots\hfill&
  \,\vdots\hfill&&\,\vdots\hfill\cr
c_n&c_{n+1}\hfill&c_{n+2}\hfill&\ldots&c_{2n}\hfill\cr
\end{matrix}\right|>0$$


Lapidus and Pindar, Numerical Solution of Partial Differential Equations in Science and
Engineering.  Page 54

$$
\int_t\left\{\sum_{j=1}^3 T_j \left({d\phi_j\over dt}+k\phi_j\right)-kT_e\right\}w_i(t)\ dt=0,
   \qquad\quad i=1,2,3.
$$

L\&P  Galerkin method weighting functions.  Page 55

$$
\sum_{j=1}^3 T_j\int_0^1\left\{{d\phi_j\over dt} + k\phi_j\right\} \phi_i\ dt
   = \int_{0}^1k\,T_e\phi_idt, \qquad i=1,2,3 $$

Another L\&P (p145)

$$
\int_{-1}^1\!\int_{-1}^1\!\int_{-1}^1 f\big(\xi,\eta,\zeta\big)
   = \sum_{k=1}^n\sum_{j=1}^n\sum_{i=1}^n w_i w_j w_k f\big( \xi,\eta,\zeta\big).
$$

Another L\&P (p126)

$$
\int_{A_e} (\,\cdot\,) dx dy = \int_{-1}^1\!\int_{-1}^1 (\,\cdot\,) \det[J] d\xi d\eta.
$$
-->

## Chemistry 101: Symbols

Chemical formulas will look best if they are not italicized. Get around math mode's automatic italicizing in LaTeX by using the argument `$\mathrm{formula here}$`, with your formula inside the curly brackets.  (Notice the use of the backticks here which enclose text that acts as code.)

So, $\mathrm{Fe_2^{2+}Cr_2O_4}$ is written `$\mathrm{Fe_2^{2+}Cr_2O_4}$`.

<!--
The \noindent command below does what you'd expect:  it forces the current line/paragraph to not indent. This was done here to match the format of the LaTeX thesis PDF.
-->

\noindent Exponent or Superscript: $\mathrm{O^-}$

\noindent Subscript: $\mathrm{CH_4}$

To stack numbers or letters as in $\mathrm{Fe_2^{2+}}$, the subscript is defined first, and then the superscript is defined.

\noindent Bullet: CuCl $\bullet$ $\mathrm{7H_{2}O}$


\noindent Delta: $\Delta$

\noindent Reaction Arrows: $\longrightarrow$ or  $\xrightarrow{solution}$

\noindent Resonance Arrows: $\leftrightarrow$

\noindent Reversible Reaction Arrows: $\rightleftharpoons$

### Typesetting reactions

You may wish to put your reaction in an equation environment, which means that LaTeX will place the reaction where it fits and will number the equations for you. 

\begin{equation}
  \mathrm{C_6H_{12}O_6  + 6O_2} \longrightarrow \mathrm{6CO_2 + 6H_2O}
  (\#eq:reaction)
\end{equation}

We can reference this combustion of glucose reaction via Equation \@ref(eq:reaction).

### Other examples of reactions

$\mathrm{NH_4Cl_{(s)}}$ $\rightleftharpoons$ $\mathrm{NH_{3(g)}+HCl_{(g)}}$

\noindent $\mathrm{MeCH_2Br + Mg}$ $\xrightarrow[below]{above}$ $\mathrm{MeCH_2\bullet Mg \bullet Br}$

## Physics

Many of the symbols you will need can be found on the math page <https://web.reed.edu/cis/help/latex/math.html> and the Comprehensive LaTeX Symbol Guide (<https://mirror.utexas.edu/ctan/info/symbols/comprehensive/symbols-letter.pdf>).

## Biology

You will probably find the resources at <https://www.lecb.ncifcrf.gov/~toms/latex.html> helpful, particularly the links to bsts for various journals. You may also be interested in TeXShade for nucleotide typesetting (<https://homepages.uni-tuebingen.de/beitz/txe.html>).  Be sure to read the proceeding chapter on graphics and tables.

----------------------------------------------------------------------------------------------

Here we will explore the climate data for the sandy site.
```{r, echo = F}
sandy_climate <- read_csv("~/Desktop/FALL_2024/Thesis/SANDY_PRISM_ppt_tmin_tmean_tmax_vpdmin_vpdmax_stable_4km_19810101_20240101_45.4727_-122.2958.csv")
sandy_climate_info <- sandy_climate %>%
  slice(1:10)
sandy_climate <- sandy_climate %>%
  slice(11:15716) %>% 
  separate(col = `PRISM Time Series Data`, c("date", "ppt", "tmin", "tmean", "tmax", "vpdmin", "vpdmax"), ",")

sandy_climate$date <- as.Date(sandy_climate$date)
```
```{r, echo = T}
plot(sandy_climate$date,sandy_climate$vpdmax, xlim = c(as.Date("1981-01-01"), as.Date("2024-01-01")))
```

###Simulating proxy anatomical time series to test using climwin

###First, here's data downloaded from test image from AutoCellRow (called it 1981)
```{r}
test <- read_csv("~/Desktop/FALL_2024/Thesis/test-quant-wood-anatomy.csv")

new <- c()
date <- c()
for (i in 1:5) {
  for (i in 1:43) {
  new <- c(new,sample(x=test[`cell type` = "Late"]$`doule wall size`, size=10, replace = TRUE))
  date <- c(date, i+1980,i+1980,i+1980,i+1980,i+1980,i+1980,i+1980,i+1980,i+1980,i+1980)
}
}
simdata <- data.frame(date, new)
simsum <- simdata %>% 
  group_by(as.factor(date)) %>%
  summarize(mean = mean(new)) %>%
  mutate(date = `as.factor(date)`) %>%
  select(date, mean)
```
##Now we have a presumably randomly simulated anatomical time series with yearly mean latewood double wall size as the biological variable that we can compare with the climate data

```{r}
#preparing sandy_climate for slidingwin formats (dd/mm/yyyy)
sandy_climate <- sandy_climate %>%
  separate(col="date", c("Y","mm","dd"),"-") %>%
  mutate(date = paste0(dd,"/",mm,"/",Y)) %>%
  select(date,ppt,tmin,tmean,tmax,vpdmin,vpdmax)
sandy_climate$vpdmax <- as.numeric(sandy_climate$vpdmax)

#preparing simsum for slidingwin format (dd/mm/yyyy)
###since we are averaging latewood growth for the year, we want to choose a date for the biological data of the end of the latewood growth (guessing it is the middle of autumn October 30th)
simdata <- simdata %>% 
  filter(date != 1981)
simdata <- simdata %>%
  mutate(Date = paste0("30/10/",date)) %>%
  select(Date,new)
```

##using climwin
```{r}
##Here i try slidingwin using vpdmax on double wall size means using a linear model and considering the past 16 months (480 days)
output <- slidingwin(xvar=list(vpdmax=sandy_climate$vpdmax),
                     cdate=sandy_climate$date,
                     bdate=simdata$Date,
                     baseline=lm(new ~ 1, data = simdata),
                     range=c(480,0),
                     type="relative",stat="mean",
                     func=c("lin"),cmissing=FALSE,cinterval="day")

dataset <- output[[1]][[3]]
plotdelta(dataset=dataset)
plotall(dataset=dataset)
```

```{r}
##Now we can randomize the dates for the biological data in order to check the probability of getting the climate response taht we got using climwin

#Creating new dataframes to hold the randomized values generated in the for loop
simdata_randomizedDate <- simdata
dataset_randomizedDate_master <- data.frame(matrix(nrow=0,ncol=length(colnames(dataset))))
colnames(dataset_randomizedDate_master) <- colnames(dataset)
best_model_randomized_master <- data.frame(matrix(nrow=0,ncol=1))
colnames(best_model_randomized_master) <- c("deltaAICc")

#will do this 100 times when it is for real (hopefully on diff computer!)
for (i in 1:100) {
  #Randomizing the date column  
  simdata_randomizedDate$Date <- sample(simdata$Date, replace = FALSE)
  #running slidingwin on randomized data
    output_randomized <- slidingwin(
      xvar=list(vpdmax=sandy_climate$vpdmax),
      cdate=sandy_climate$date,
      bdate=simdata_randomizedDate$Date,
      baseline=lm(new ~ 1, data = simdata_randomizedDate),
      range=c(480,0),
      type="relative",stat="mean",
      func=c("lin"),cmissing=FALSE,cinterval="day")
    #Extracting the dataset with deltaAICc values
    dataset_randomized <- output_randomized[[1]][[3]]
    #adding the best model to the master list
    best_model_randomized_master <- rbind(best_model_randomized_master,dataset_randomized$deltaAICc[1])
    #Adding these values to the master dataframe that will be used to find probabilities
    dataset_randomizedDate_master <- rbind(dataset_randomizedDate_master,dataset_randomized)
}

#Visualizing the distribution of deltaAICc values
hist(dataset_randomizedDate_master$deltaAICc)
hist(best_model_randomized_master$deltaAICc)

#Here I will show an example of calculated the probability of getting a best model with a deltaAICc value that we get or lower using the randomized master list of deltaAICcs we just made and plotted
#First I will create a dataset of of 100 values from rnorm to serve as the best_model_randomized_master$deltaAICc distribution
distribution <- rnorm(1000, mean = 0, sd = 1)
#Then I create an empirical cumulative distribution function from that sampled distribution
ecdf_func <- ecdf(distribution)
#Finally this will calculate the probability that we got the deltaAICc value or lower (in this case it is just -2) 
prob <- ecdf_func(-2)
```






